---
title: "Reproducible Research: Peer Assessment 1"
author: "David Dessert"
date: "Tuesday, October 14, 2014"
output: html_document
keep_md: true
---

## Load necessary packages to execute
```{r}
library(reshape2)
library(ggplot2)
library(scales)
library(zoo)
```

## Load and preprocessing the data
The input data file is stored in the ./data subdirectory and
is in a comma-separated variable format (.csv extension).
The first row contains the header names {steps, date, interval}.
```{r loading}
inFile <- './data/activity.csv'
df_activity <- read.csv(inFile)
head(df_activity)
```
A time field is computed from the "date" and "interval".  
"steps" that are NA are identified.
```{r processing}
# Calculate POSIX time from date and interval fields
df_activity$time <- as.POSIXct(3600*floor(df_activity$interval/100)
                               + 60*(df_activity$interval %% 100), 
                               tz = "GMT", 
                               origin = as.POSIXct(df_activity$date, tz = "GMT"))
# Mark all NA fields
df_activity$na_steps <- is.na(df_activity$steps)
# Minutes within the day
df_activity$minutes <- 60*floor(df_activity$interval/100) +
                       (df_activity$interval %% 100)
# Number of days within the record
NUM_DAYS <- length(unique(df_activity$date))
# Number of time intervals within a day
NUM_INTERVALS_PER_DAY <- length(unique(df_activity$interval))
head(df_activity)
```


## Find the total number of steps taken per day
```{r total_steps_per_day}
# Calculate the total # steps for each day
daily_steps <- aggregate(steps ~ date, FUN = sum, data = df_activity)
head(daily_steps)

# Histogram plot
p <- qplot(steps, data = daily_steps, geom = "histogram", binwidth = 1000)
p <- p + xlab("Daily Steps")                    # x-axis label
p <- p + ylab("Count")                          # y-axis label
p <- p + ggtitle("Histogram of Daily Steps")    # graph title
print(p)
```

```{r}
# Calculate the mean and median of the original data
mean_daily_steps <- sprintf("%.1f", mean(daily_steps[,2]))
median_daily_steps <- sprintf("%d", median(daily_steps[,2]))
```
The mean steps per day is `r mean_daily_steps`.  
The median steps per day is `r median_daily_steps`.

## Display the average daily activity pattern
Create a new data frame holding the 5-minute time interval and the sum of all steps from all days at that specific time interval.
```{r daily_activity}
time_steps <- aggregate(steps ~ minutes, FUN = sum, data = df_activity)
head(time_steps)
```
```{r plot_daily_activity}
# Find the maximum step time and value
i_max <- which.max(time_steps[,2])
t_max <- sprintf("%2d:%02d", 
                 floor(time_steps[i_max,1] / 60),
                 time_steps[i_max,1] %% 60)
s_max <- time_steps[i_max,2] / NUM_DAYS

# Plot the averages over time
p <- ggplot(time_steps, aes(x=minutes, y=steps/NUM_DAYS))
# Line graph
p <- p + geom_line(size=1)
# Red point for the maximum step time and value
p <- p + annotate("point", x=time_steps$minutes[i_max], 
                           y=s_max, 
                           colour="red")
# Text indicating the maximum step time and value
p <- p + annotate("text", 
                  x=time_steps$minutes[i_max], 
                  y=s_max, 
                  hjust=0,
                  label=sprintf("  (%s, %.1f)", t_max, s_max),
                  colour = "red")
# Pretty up the x-axis
# Calculate 5 even breaks on x-axis
t_breaks <- seq(from=1, to=NUM_INTERVALS_PER_DAY, length.out=5);
# Add the breaks and labels (that don't have the day information)
p <- p + scale_x_continuous(breaks = time_steps$minutes[t_breaks],
                            labels = c("00:00","06:00","12:00","18:00","24:00"))
p <- p + xlab("Time of Day")                    # x-axis label
p <- p + ylab("Steps")                          # y-axis label
p <- p + ggtitle("Mean Steps by Time of Day")   # graph title
print(p)
```
  
The maximum average daily steps is `r s_max`, occured at time `r t_max` and is drawn in red.


## Imputing missing values
Missing (NA) values with non-missing adjacent step values are linearly interpolated in
time to impute their new value. After this, the first and last day's step values 
remained missing. Their missing step values are imputed by using the time-of-day
adjacent step values.
```{r impute_missing_values}
# Count of NA in original step data
num_na <- sum(df_activity$na_steps)
# Use the zoo package to calculate interpolated NA values
# Start with the original step data
df_activity$steps_interp <- df_activity$steps
# Fill in NA (imputed values)
# Interpolate steps using the time of day (interval) value
zz <- round(na.approx(df_activity$steps_interp, df_activity$interval, na.rm = FALSE))
# Replace only the remaining NA values
df_activity$steps_interp[is.na(df_activity$steps_interp)] <- zz[is.na(df_activity$steps_interp)]

## 
matrix_time_by_steps_interp <- matrix(df_activity$steps_interp, nrow = 24*12, byrow = FALSE)
# Total Steps for each day
daily_steps_interp <- colSums(matrix_time_by_steps_interp)
```

```{r plot_imputed_values}
# Histogram plot
p <- ggplot(data.frame(x=daily_steps_interp, y=1:length(daily_steps_interp)), aes(x)) 
p <- p + geom_histogram(binwidth = 1000)
p <- p + xlab("Daily Steps")                    # x-axis label
p <- p + ylab("Count")                          # y-axis label
p <- p + ggtitle(expression(atop("Histogram of Daily Steps",
                                 "NA's Interpolated")))     # graph title
print(p)
```

```{r}
# Calculate the mean and median of the imputed data
mean_daily_steps_interp <- sprintf("%.1f", mean(daily_steps_interp, na.rm = TRUE))
median_daily_steps_interp <- sprintf("%d", median(daily_steps_interp, na.rm = TRUE))
```
  
In the original data set, there are `r num_na` missing step values (indicated with NA).
After imputing the missing (NA) values, the mean steps per day is 
`r mean_daily_steps_interp` versus `r mean_daily_steps` (raw data).  
After imputing the missing (NA) values, the median steps per day is 
`r median_daily_steps_interp` versus `r median_daily_steps` (raw data).  

The following graph shows the original data steps and the newly imputed steps.
```{r plot_orig_and_imputed}
m <- melt(df_activity, 
          id.vars = "time", 
          measure.vars = c("steps_interp","steps"),
          variable.name = "type",
          value.name = "steps")
ggplot() + geom_point(data=m, aes(x=time, y=steps, group=type, colour=type))
```

## Show the differences in activity patterns between weekdays and weekends
```{r identify_weekday_weekend_diffs}
# Identify the weekdays and weekends
df_activity$weekday <- factor(weekdays(df_activity$time) %in% c('Saturday','Sunday'),
                              levels=c("TRUE", "FALSE"), 
                              labels=c("Weekend","Weekday"))


# Calculate the mean # steps for each weekday/interval pair
# steps_by_day_interval <- aggregate(steps ~ interval + weekday, FUN = mean, data = m)
steps_by_day_interval <- aggregate(steps ~ minutes + weekday, FUN = mean, data = df_activity)

# Calculate the interval values for each 6-hour period
t_breaks <- seq(1, max(df_activity$minutes)+5, length.out=5)
```

```{r plot_weekday_weekend_diffs}
# Plot the weekend & weekday mean steps throughout the day
p <- ggplot(steps_by_day_interval, aes(minutes, steps))
p <- p + geom_line()                            # line plot
p <- p + facet_grid(weekday ~ .)                # Top/Bottom plot
p <- p + scale_x_continuous(breaks = df_activity$minutes[t_breaks],
                            labels = c("00:00","06:00","12:00","18:00","24:00"))
p <- p + xlab("Time of Day")                    # x-axis label
p <- p + ylab("Steps")                          # y-axis label
p <- p + ggtitle("Mean Steps by Time of Day")   # graph title
print(p)
```